Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-04 tagless/tagless-final.tex
Done with Section 4

\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

\begin{abstract}
We have built a family of tagless interpretations for a
higher-order typed object
language in a typed metalanguage (Haskell and ML), 
requiring neither dependent
types, nor generalized algebraic data types, nor a tag elimination 
postprocessing step. That has been an open problem.
Our family of interpreters contains 
an evaluator, a (byte-)compiler and a staged evaluator, a partial
evaluator, a term-size evaluator and a CPS call-by-name and
call-by-value transformers.

@@

We focus on embedding a typed object language into a typed metalanguage.
The particular benefit is ruling out meaningless object terms, and
so enabling faster interpretation. In addition, our interpreters
assuredly do not get stuck.
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just for
evaluation of object programs but also for
compilation, partial evaluation, and other processing.

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  
No matter how we represent the object language in the
metalanguage, the representation can be performed by hand (e.g., by
entering object terms at a metalanguage interpreter's prompt) or
created by a parser/typechecker reading from a text string.
We leave aside
the issue of whether to program the parser/typechecker using dependent
types \citep{WalidICFP02} or not \citep{baars-typing}.
\oleg{perhaps say more affirmatively: the problem of writing such
  a parser/interpreter has been solved, with dependent types (cite) and
  without (cite), and so will not be discussed here.}

\subsection{The Tag Problem}\label{tagproblem}

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp
\end{code}
We represent each variable using an unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

@@

\subsection{Solutions Using Fancier Types}

@@

\subsection{Our Final Proposal}\label{ourapproach}

@@

Our contributions are
\begin{enumerate}
\item to evaluate, compile, and partially evaluate a typed object language
   in a typed metalanguage with a simple type system;
\item to check statically for well-typed and closed object programs;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CBN CPS;
\item a clear way to extend the object language with more features;
\item and an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out to be harder than expected.
\end{enumerate}

@@

The structure of our paper is as follows. In the next Section, we
define the OCaml and Haskell interfaces of the interpreter for our
language and describe two instances, evaluating a term to a
metalanguage value and computing term's size. In
Section~\ref{compiler}, we stage the evaluator, obtaining the tagless
compiler. Section~\ref{PE} describes partial evaluation, in
particular, the problem posed by the need to delay the 
binding-time analysis of term abstractions and the apparent 
need for a typecase. We then show two solutions. The first, imperfect
one, uses GADT. The better and more portable solution bakes in the
typecase idiom into the interpreter interface, eliminating the need for
advanced type systems and permitting implementations in Haskell98 or OCaml.
The discussion Section~\ref{discussion} presents, inter alia, CBN CPS
interpreter and the CBV CPS transformer. We then describe 
self-interpretation (Section~\label{selfinterp}),
discuss related work (Section~\ref{related}) and then conclude.

@@

The functor below encodes our examples |test1| and $\mathit{power}$.
\begin{code}
module EX(S: Symantics) = struct
  open S
  let test1 () =
    app (lam (fun x -> x)) (bool true)
  let testpowfix () = 
    lam (fun x -> fix (fun self -> lam (fun n ->
    if_ (leq n (int 0)) (fun () -> int 1)
        (fun () -> mul x
          (app self (add n (int (-1))))))))
  let testpowfix7 = 
     lam (fun x -> app (app (testpowfix ()) x)
                       (int 7))
end;;
\end{code}
The dummy argument to |test1| and |testpowfix| is an artefact of
MetaOCaml, again related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction then dictates that
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

end
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler simply produces
residual code with no optimization. For example, interpreting our |test1|
\begin{code}
let module E = EX(C) in E.test1 ()
\end{code}
gives the code value |.<(fun x_6 -> x_6) (true)>.|
of inferred type |('c, bool) C.repr|.

@@

\ccshan{Move later:} The
partial evaluator does have tags: \emph{binding time} or \emph{phase}
tags, i.e. whether a value is static or dynamic.
Pattern-matching on this phase tag is always exhaustive.

@@

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

@@

end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (eliding \texttt{mul} and \texttt{leq})}
\label{fig:pe}
\end{figure}

@@

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| literally expresses the type equation for |repr_pe| above.
The function |abstr|, as in~\S\ref{S:PE-GADT},
extracts a future-stage code value from a result of
partial evaluation.  Conversely, the function |pdyn| coerces a
code value into a partial\hyp evaluation result. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
static ones, if any, to avoid
polymorphic lift. The implementations of
|add| and |mul| are a bit more sophisticated, taking advantage 
respectively of the implicit monoid and ring structure of the 
underlying domain.
The function |build| checks to see if
its two arguments |e1| and |e2| are static expressions. If so, the
corresponding static computation (with the help of the intrepreter
|R|) is performed. The function |buildsim_monoid| also checks if one argument
is static and is the neutral element of the monoid structure.
|ring| also worries about the full ring structure rather than
just the monoid structure.  These are put here to illustrate how we 
would go about writing general algebraic simplification routines that
would mimic various compiler optimizations, and these would be quite
beneficial in a large language with more base types.

When doing partial evaluation on |fix|, we made a deliberate decision to `go
all the way' -- so long as the computation can proceed at the PE stage,
we do that rather than residualizing. In the
accompanying source code, we have also implemented an alternative
where |fix| is unrolled only once and then we residualize. We prefer the
first approach, even though it may cause
non-termination at the PE stage -- but \emph{only} if the computation
we are partially evaluating is non-terminating, and enough static
information is available to trigger this non-termination.  Our
treatment of |fix| may lead to code bloat in the residual
program. The techniques of~\citet{SwadiTahaKiselyovPasalic2006} can
be used to deal with this issue.

Given the above Symantics, our running test now evaluates
\begin{code}
let module E = EX(P) in E.test1 ();;
\end{code}
to
\begin{code}
- : ('a, bool, bool) P.repr = 
   {P.st = Some true; P.dy = .<(true)>.}
\end{code}
It is instructive to compare this result with that of the |C|
evaluator. We observe that the beta-redex has now been 
performed. The result is the static value |true|.
More interesting is the partial evaluation of testpowfix7. The |C|
interpreter evaluates |testpowfix7| to the future-stage code with many
apparent beta-redexes.
\begin{code}
let module E = EX(C) in E.testpowfix7;;
 - : ('_a, int -> int) C.repr =
.<fun x_1 -> (((fun x_2 ->
  let rec self_3 =
   fun n_4 ->
    ((fun x_5 -> if (x_5 <= 0) then 1 
                 else (x_2 * (self_3 (x_5 + (-1)))))
      n_4) in
  self_3) x_1) 7)>.
\end{code}
The result of the |P| interpreter is different:

\begin{code}
let module E = EX(P) in E.testpowfix7;;
 - : ('_a, ('_a, int, int) P.repr -> ('_a, int, int) P.repr, 
      int -> int) P.repr = 
 {P.st = Some <fun>;
  P.dy = .<fun x_1 -> 
     (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1))))))>.}
\end{code}
as expected from the power example.

Unlike the GADT approach in the previous subsection, there is no
longer any seemingly partial pattern-matching. All pattern-matching
occurs in the partial evaluator only (to check if the current value 
is static or dynamic), and is 
phase rather than a type tag match. Furthermore the matching
\emph{patently} exhaustive. That is, totality is assured and it is
assured \emph{syntactically}, in a way that is apparent to the
compiler.

One should note that the partial evaluator has fully reused the code
of the compiler and the interpreter. It is truly the `composition' of the
interpreter and the compiler.   This is a simpler situation than
in \citet{SperberThiemann:TwoForOne} where a partial evaluator and
a compiler are `composed', but the general ideas are similar.

Lastly, we should comment on the approach taken in the partial evaluator.
It is an online, typed, polyvariant partial evaluator.  It has no
termination analysis, and as written may diverge (but only on divergent
programs) as we have decided to aggressively unroll loops.

@@

\subsection{Lifting and Functor}

>>>>

\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}
\usepackage{dblfloatfix}

@@

\begin{abstract}
We have built a family of tagless interpretations for a
higher-order typed object
language in a typed metalanguage (Haskell and ML), 
requiring neither dependent
types, nor generalized algebraic data types, nor a tag elimination 
postprocessing step. That has been an open problem.
Our family of interpreters contains 
an evaluator, a (byte-)compiler and staged evaluator, a partial
evaluator, a term-size evaluator, and call-by-name and
call-by-value CPS transformers.

@@

We focus on embedding a typed object language into a typed metalanguage.
The benefit of types in this setting is to rule out meaningless object terms,
thus enabling faster interpretation and assuring that our interpreters
do not get stuck.
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just for
evaluation of object programs but also for
compilation, partial evaluation, and other processing.

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  
No matter how we represent the object language in the
metalanguage, the representation can be created either by hand (for example, by
entering object terms at a metalanguage interpreter's prompt) or
by a parser or type checker reading from a text string.
We leave aside the solved problem of writing such a parser or type checker,
whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.

\subsection{The tag problem}\label{tagproblem}

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp
\end{code}
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

@@

\subsection{Solutions using fancier types}

@@

\subsection{Our final proposal}\label{ourapproach}

@@

Our contributions are
\begin{enumerate}
\item to evaluate, compile, and partially evaluate a typed object language
   in a typed metalanguage with a simple type system;
\item to check statically for well-typed and closed object programs;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CPS transformation;
\item a clear way to extend the object language with more features;
\item and an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out to be harder than expected.
\end{enumerate}

@@

The structure of our paper is as follows. In the next Section, we
define the OCaml and Haskell interfaces of the interpreter for our
language and describe two instances, evaluating a term to a
metalanguage value and computing term's size. In
Section~\ref{compiler}, we stage the evaluator, obtaining the tagless
compiler. Section~\ref{PE} describes partial evaluation, in
particular, the problem posed by the need to delay the 
binding-time analysis of term abstractions and the apparent 
need for a typecase. We then show two solutions. The first, imperfect
one, uses GADT. The better and more portable solution bakes in the
typecase idiom into the interpreter interface, eliminating the need for
advanced type systems and permitting implementations in Haskell98 or OCaml.
The discussion Section~\ref{discussion} presents, inter alia, CBN CPS
interpreter and the CBV CPS transformer. We then describe 
self-interpretation (Section~\ref{selfinterp}),
discuss related work (Section~\ref{related}) and then conclude.

@@

The functor below encodes our running examples |test1| and $\mathit{power}$.
\begin{code}
module EX(S: Symantics) = struct
  open S
  let test1 () =
    app (lam (fun x -> x)) (bool true)
  let testpowfix () = 
    lam (fun x -> fix (fun self -> lam (fun n ->
    if_ (leq n (int 0)) (fun () -> int 1)
        (fun () -> mul x
          (app self (add n (int (-1))))))))
  let testpowfix7 = 
     lam (fun x -> app (app (testpowfix ()) x)
                       (int 7))
end;;
\end{code}
The dummy argument to |test1| and |testpowfix| is an artefact of
MetaOCaml, again related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction then dictates that
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

end
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler simply produces
residual code with no optimization. For example, interpreting our |test1|
\begin{code}
let module E = EX(C) in E.test1 ()
\end{code}
gives the code value |.<(fun x_6 -> x_6) (true)>.|
of inferred type |('c, bool) C.repr|.  Similarly, interpreting |testpowfix7|
with
\begin{code}
let module E = EX(C) in E.testpowfix7
\end{code}
gives the code value
\begin{code}
.<fun x_1 -> (fun x_2 ->
  let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1 
               else x_2 * (self_3 (x_5 + (-1))))
   n_4
  in self_3) x_1 7>.
\end{code}
with many apparent $\beta$-redexes.

@@

@@

@@

end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
\label{fig:pe}
\end{figure}

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.

@@

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| literally expresses the type equation for |repr_pe| above.
The function |abstr|, as in~\S\ref{S:PE-GADT},
extracts a future-stage code value from a result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into partial evaluation. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to avoid polymorphic lift.

To illustrate how to add optimizations, we improve |add| (and |mul|,
elided) to simplify the generated code using the monoid (and ring)
structure of~|int|: not only is addition performed statically
(using~|R|) when both operands are statically known, but it is
eliminated when one operand is statically~$0$; similarly for
multiplication by~$0$ or~$1$.  Such algebraic simplifications are easy
to abstract over the specific domain (such as monoid or ring) where they
apply.  These simplifications and abstractions help a lot in a large
language with more base types and primitive operations.

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, unfold |fix| rather than residualize whenever the
argument is static.  In the accompanying source code is a conservative
alternative |P.fix| that unrolls recursion only once, then residualizes.
Many sophisticated approaches have been developed to make this trade-off
\citep{jones-partial}, but this issue is orthogonal to our presentation.
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{SwadiTahaKiselyovPasalic2006}.

Given this implementation of~|P|, our running example
\begin{code}
let module E = EX(P) in E.test1 ()
\end{code}
evaluates to
\begin{code}
{P.st = Some true; P.dy = .<true>.}
\end{code}
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
\begin{code}
let module E = EX(P) in E.testpowfix7
\end{code}
gives
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 *
                    (x_1 * (x_1 * x_1)))))>.}
\end{code}
as desired.

Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp matching
in~|P| is \emph{syntactically} exhaustive, so it is \emph{patent} to the
compiler that |P| is total.  Further, all pattern\hyp matching occurs
during partial evaluation, only to check if a value is known statically,
never what type it has.  In other words, our partial evaluator tags
phases (with |Some| and |None|) but not object types.

Our typed partial evaluator is online and polyvariant.  It reuses the
compiler~|C| and the evaluator~|R| by composing them.  This situation is
simpler than \citets{SperberThiemann:TwoForOne} composition of a partial
evaluator and a compiler, but the general ideas are similar.

@@

\subsection{Lifting and functor}
